Life in the canopy: community trait assessments reveal substantial functional diversity among fern epiphytes

### 2019-10-27 ms/manuscript.Rmd
Update MS
Finalize title and authors, summary, add Fig S2

---
title: "Constrained, yet diverse: Phylogenetic and morphological filtering in epiphytic ferns"
bibliography: "references.bib"
mainfont: Times New Roman.ttf
fontsize: 12pt
mainfontoptions: 
- BoldFont=Times New Roman Bold.ttf
- ItalicFont=Times New Roman Italic.ttf
- BoldItalicFont=Times New Roman Bold Italic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: none # Use pandoc default, otherwise csl can't be used for reference formatting
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "new-phytologist.csl"]
    includes:
      in_header: newphyt.sty
header-includes: 
  \usepackage{float}
  \makeatletter\renewcommand*{\fps@figure}{H}\makeatother
editor_options: 
  chunk_output_type: console
nocite: | 
  @Vogel1968
---

@@

Joel H. Nitta^1,4^, James E. Watkins, Jr.^2^, Ravahere Taputuarai^3^, and Charles C. Davis^1^

@@

^3^Association Te Rau Ati Ati a Tau a Hiti Noa Tu, B.P. 11553, Mahina, Tahiti, French Polynesia

^4^Current affiliation: Department of Botany, National Museum of Natural History, Smithsonian Institute, Washington, D.C., 20013, U.S.A.

Author for correspondence:  
*Joel H. Nitta*  
*Tel: +1 6174960515*  
*Email: joelnitta@gmail.com*

@@

- The appearance of angiosperm-dominated forests in the Cretaceous and early Cenozoic had a profound effect on terrestrial biota by creating novel ecological niches. The diversity of modern fern lineages is thought to be a response to angiosperm dominance, radiating both in the canopy and on the forest floor. Despite the significance of epiphytic growth for fern evolution, however, few studies have investigated traits related to epiphytism in ferns while accounting for phylogeny.

- We investigated evolutionary patterns in traits related to epiphytic growth at the species and community level using a phylogeny including representatives of multiple epiphytic fern radiations on the island of Moorea, French Polynesia.

- We observed phylogenetic conservatism in multiple traits related to epiphytic growth in fern sporophytes and gametophytes. Epiphytes had significantly shorter fronds, smaller rhizomes, and a higher frequency of non-cordate gametophytes. Epiphytic communities were clustered phylogenetically but diverse functionally, indicating that the canopy acts as a filter due to its low humidity to limit size while simultaneously enabling diversification into finely divided niches.

- Our results provide an important phylogenetic perspective on the role of functional traits in ferns, and pave the way for more detailed studies examining how niche space is filled within terrestrial and epiphytic habitats.

@@

>>>>

---
title: "Life in the canopy: community trait assessments reveal substantial functional diversity among fern epiphytes"
bibliography: "references.bib"
mainfont: Times New Roman.ttf
fontsize: 12pt
mainfontoptions: 
- BoldFont=Times New Roman Bold.ttf
- ItalicFont=Times New Roman Italic.ttf
- BoldItalicFont=Times New Roman Bold Italic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: none # Use pandoc default, otherwise csl can't be used for reference formatting
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: no
    latex_engine: xelatex
    pandoc_args: [ "--csl", "new-phytologist.csl"]
    includes:
      in_header: newphyt.sty
header-includes: 
  \usepackage{float}
  \makeatletter\renewcommand*{\fps@figure}{H}\makeatother
editor_options: 
  chunk_output_type: console
nocite: | 
  @Vogel1968
---

@@

Joel H. Nitta^1,3^, James E. Watkins, Jr.^2^, and Charles C. Davis^1^

@@

^3^Current affiliation: Department of Botany, National Museum of Natural History, Smithsonian Institution, Washington, D.C., 20013, U.S.A.

Authors for correspondence:  
*Joel H. Nitta*  
*Tel: +1 2026331000*  
*Email: joelnitta@gmail.com*

*Charles C. Davis*  
*Tel: +1 6174960515*  
*Email: cdavis@oeb.harvard.edu*

@@

- The expansion of angiosperm-dominated forests in the Cretaceous and early Cenozoic had a profound effect on terrestrial biota by creating novel ecological niches. The majority of modern fern lineages are hypothesized to have arisen in response to this expansion, particularly fern epiphytes that radiated into the canopy. Recent evidence, however, suggests that epiphytism does not correlate with increased diversification rates in ferns, calling into question the role of the canopy habitat in their diversification.

- To bridge this impasse, we investigated functional traits of fern sporophytes and gametophytes at the species and community level across a broad phylogenetic sampling on the island of Moorea, French Polynesia, including > 120 species and representatives of five major epiphytic radiations.

- While epiphytes showed convergence in small size and a higher frequency of non-cordate gametophytes, they showed greater functional diversity at the community level relative to terrestrial ferns.

- These results suggest previously overlooked functional diversity among fern epiphytes, and raise the hypothesis that while the angiosperm canopy acted as a complex filter that restricted plant size, it also facilitated diversification into finely partitioned niches. Characterizing these niche axes and adaptations of epiphytic ferns occupying them should be a priority for future pteridological research.

@@

**Fig. S2** Comparison of community-weighted standard deviations (CWSD) in traits of epiphytic and terrestrial fern communities on Moorea, French Polynesia.
